The Talking Heads Experiment

### 2015-04-10 chap4/chap4.tex
cumulative update, mainly for Ch. 7-12
minor changes

Early rationalists, like Descartes, were dualists, which
saw the mind and the body in totally different realms. 
In such a view, it becomes very difficult to scientifically investigate 
perceptually grounded categorisation, and an explanation 
how the brain works seems more remote than ever. However, 
most contemporary rationalists (like empiricists)
now believe that categorisation
is done by physical structures in the brain. Indeed, 
if certain parts of the brain are damaged, the capability 
to categorise disappears or is severely restricted and 
distorted (\cite{Deacon:1998}).

Rationalists claim that categories 
exist {\itshape a priori} and therefore categorisation comes from
within. They argue that humans have a repertoire of ideal universal
forms, which they project
onto reality. Reality itself is a weak, imperfect 
reflection of these forms, like the shadows of objects
on the wall of a cave. Because of this poverty of stimuli, 
categories (particularly the perceptually grounded
categories that are the focus of our attention in this 
book) are claimed to be unlearnable by induction and must
therefore be innate.\footnote{
Strong forms of innateness have been defended by 
\cite{Fodor:1983}
and many philosophers and linguists associated with the 
generative grammar paradigm. See the discussion in
\cite{Wierzbicka:1992}, particularly the introductory chapter. 
In artificial intelligence 
research, particularly the logic-oriented tradition, there
is often an implicit acceptance that basic categories 
are innate, but that derived categories, formulatable
in terms of more primitive concepts, can be learned
(see \cite{McCarthy:2008}). There are also 
large-scale efforts going on to build an ontology as 
rich as human ontologies, see the discussions around
Doug Lenat's CYC project in: \cite{Steels:1994}.}

@@

In the case of the immune system, a solution is an 
antibody capable of combatting intruders foreign to the 
organism. Here again an instructionist approach can 
be envisioned and was believed for a long time to be 
the case. For this belief to be true would mean that
the immune system somehow learns the appropriate response
and then preserves that response. The selectionist viewpoint
of the immune system argues instead that it
generates autonomously 
a very large repertoire of possible antibody responses. 
When a foreign body invades, the response 
is already 
there, it is simply amplified (\cite{Varela:1988}).

@@

The growth of a tree or plant is a good 
metaphor to visualise this selectionist approach. 
The shape of a tree appears well adapted to 
its environment. Typically there are more branches and leaves 
where there is more sunlight. The height of
the tree reflects the competition of neighbouring trees
or the height of surrounding buildings. The overall shape reflects
the shape of surrounding walls or other trees. 
It is obvious that a tree
does not come with "shape genes" that determine
exactly which shape the tree will have in a particular 
setting. Nor does it come with sophisticated sensors
and a brain inductively learning about the environment
so as to decide on which branch the next leaf should grow. 
Instead the tree grows in all directions following
a steady, usually quite regular growth pattern. A tree
standing alone in a landscape exhibits a beautiful 
balanced shape, expanding in all directions, but when the 
growth is constrained, the tree reflects these constraints. 
The branches and leaves that catch more sunlight receive more resources 
to flourish and develop further, whereas those 
pointing towards an area with no sunlight are stifled in 
their growth and may die altogether.

>>>>

Early rationalists, like Descartes, were dualists, which
saw the mind and the body in totally different realms. 
In such a view, it becomes very difficult to scientifically investigate 
perceptually grounded categorisation, and an explanation 
how the brain works seems more remote than ever. However, 
most contemporary rationalists (like empiricists)
now believe that categorisation
is done by physical structures in the brain. Indeed, 
if certain parts of the brain are damaged, the capability 
to categorise disappears or is severely restricted and 
distorted (\citealt{Deacon:1998}).

Rationalists claim that categories 
exist {\itshape a priori} and therefore categorisation comes from
within. They argue that humans have a repertoire of ideal universal
forms, which they project
onto reality. Reality itself is a weak, imperfect 
reflection of these forms, like the shadows of objects
on the wall of a cave. Because of this poverty of stimuli, 
categories (particularly the perceptually grounded
categories that are the focus of our attention in this 
book) are claimed to be unlearnable by induction and must
therefore be innate.\footnote{
Strong forms of innateness have been defended by 
\cite{Fodor:1983}
and many philosophers and linguists associated with the 
generative grammar paradigm. See the discussion in
\cite{Wierzbicka:1992}, particularly the introductory chapter. 
In artificial intelligence 
research, particularly the logic-oriented tradition, there
is often an implicit acceptance that basic categories 
are innate, but that derived categories, formulatable
in terms of more primitive concepts, can be learned
(see \citealt{McCarthy:2008}). There are also 
large-scale efforts going on to build an ontology as 
rich as human ontologies, see the discussions around
Doug Lenat's CYC project in: \cite{Steels:1994}.}

@@

In the case of the immune system, a solution is an 
antibody capable of combatting intruders foreign to the 
organism. Here again an instructionist approach can 
be envisioned and was believed for a long time to be 
the case. For this belief to be true would mean that
the immune system somehow learns the appropriate response
and then preserves that response. The selectionist viewpoint
of the immune system argues instead that it
generates autonomously 
a very large repertoire of possible antibody responses. 
When a foreign body invades, the response 
is already 
there, it is simply amplified (\citealt{Varela:1988}).

@@

The growth of a tree or plant is a good 
metaphor to visualise this selectionist approach. 
The shape of a tree appears well adapted to 
its environment. Typically there are more branches and leaves 
where there is more sunlight. The height of
the tree reflects the competition of neighbouring trees
or the height of surrounding buildings. The overall shape reflects
the shape of surrounding walls or other trees. 
It is obvious that a tree
does not come with ``shape genes" that determine
exactly which shape the tree will have in a particular 
setting. Nor does it come with sophisticated sensors
and a brain inductively learning about the environment
so as to decide on which branch the next leaf should grow. 
Instead the tree grows in all directions following
a steady, usually quite regular growth pattern. A tree
standing alone in a landscape exhibits a beautiful 
balanced shape, expanding in all directions, but when the 
growth is constrained, the tree reflects these constraints. 
The branches and leaves that catch more sunlight receive more resources 
to flourish and develop further, whereas those 
pointing towards an area with no sunlight are stifled in 
their growth and may die altogether.
